---
title: "Final"
author: '[Your first name] [Your last name]'
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
---

# Instructions

* **Complete the problem sets below.**
*  **Write all of your code in the .Rmd file, with zero dependencies on external files other than the `final_data` object that I created, and that you will load in from a GitHub URL.**
* **Use your best judgment on the number of significant figures to include in your reported results. Don't just report the default. Think about the context and any comparisons you are making, as well as the scale of the measurements, and then choose an appropriate display precision and unit of measurement.**
* **Write your code, display your results, and interpret results in the areas indicated in the instructions of each problem.**
* **Do not just print out the `data.frame` of results. Put your results into a formatted table using `kable` or a similar package.**
* **Read [about the data](#about-the-data) before starting the problems to understand what data is available to you.**

# Spoiler alert

**You will not have to construct any (parent single decrement) life tables during this final. We've been there and done that.** 

# Setup

```{r}
# Install pacman
if (!require("pacman")) install.packages("pacman")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(dplyr, ggplot2, knitr, kableExtra, readr, stringr)
```

# Data collection

```{r}
final_data <- readRDS(url("https://github.com/hanowell/uwsoc533a-final/blob/main/data/final_data.rds?raw=true"))
```


# About the data

## Context

**All of the data used for the final applies to periods or cohorts in the U.S.**

## Data sources

**Data sets providing exposures, population counts, and mortality statistics come from the Human Mortality Database (HMD). Data sets providing fertility statistics come from the Human Fertility Database (HFD). Death counts used to infer the age-specific Sars-CoV-2 mortality rate comes from the National Center for Health Statistics, Data from “Deaths involving coronavirus disease 2019 (COVID-19) by race and Hispanic origin group and age, by state.”^[https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Age-in-Years-/3apk-4u4f/data]. Data used to infer age-specific fertility rates for 2020 come from **

## Top-level data structure

**You've just loaded a `list` of all the `data.frame`s you need to complete this final. To access a given `data.frame` in the `list`:**

```{r eval=FALSE}
final_data[["[name of data.frame]"]]
```

**For example:**

```{r}
head(final_data[["us_asmr"]])
```

## Data dictionaries

### `us_asfr`

**Annual period age-specific fertility rates (ASFRs) for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which ASFR was measured**
* **`Age`: Starting age of one-year age group applicable to `ASFR` in `Year`**
* **`ASFR`: ASFR for the `Age` group in `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the left or right**

### `us_asmr`

**Annual period age-specific mortality rates (ASMRs) by sex for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which ASMR was measured**
* **`Age`: Starting age of one-year age group applicable to ASMR in `Year`**
* **`Female`: Female ASMR for the `Age` group in `Year`**
* **`Male`: Male ASMR for the `Age` group in `Year`**
* **`Total`: ASMR for the `Age` group in `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the right**

### `us_births`

**Annual period age-specific birth counts on January 1st  for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which births were counted on 1 January**
* **`Age`: Starting age of one-year age group applicable to birth count in `Year`**
* **`Total`: January 1st birth count for the `Age` group in `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the left or right**

### `us_deaths`

**Annual period age-specific death counts for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which death count was measured**
* **`Age`: Starting age of one-year age group applicable to death count in `Year`**
* **`Female`: Female death count for the `Age` group in `Year`**
* **`Male`: Male death count for the `Age` group in `Year`**
* **`Total`: Total death count for the `Age` group in `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the right**

### `us_deaths_despair`

**Annual period age-specific death counts due to "despair" (alcoholic liver disease, alcohol abuse, drug abuse, and suicide or self-inflicted injury) in the U.S. in 2015.**

* **`Age`: Starting age for the age group**
* **`D_despair`: Number of deaths in `Age` group due to despair**

### `us_deaths_lexis`

**Annual period age-specific death counts by birth cohort and sex in the U.S. in 2015.**

* **`Age`: Starting age for the age group**
* **`Cohort`: Birth cohort to which age-specific death count applies**
* **`Female`: Female cohort- and age-specific death count**
* **`Male`: Male cohort- and age-specific death count**
* **`Total`: Total cohort- and age-specific death count**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the right**

### `us_exposures`

**Annual period age-specific person-year (ASPY) approximations by sex for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which ASPY was measured**
* **`Age`: Starting age of one-year age group applicable to ASPY in `Year`**
* **`Female`: Female ASPY for the `Age` group in `Year`**
* **`Male`: Male ASPY for the `Age` group in `Year`**
* **`Total`: Total ASPY for the `Age` group in `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the right**

### `us_population`

**Age-specific population counts by sex and day of year (January 1st vs. December 31st) for the U.S. from 1905 through 2016.**

* **`Year`: The one-year period during which population counts were measured**
* **`Age`: Starting age of one-year age group applicable to ASPY in `Year`**
* **`Female1`: Female population count for the `Age` group on January 1st of `Year`**
* **`Male1`: Male population count for the `Age` group on January 1st of `Year`**
* **`Total1`: Total population count for the `Age` group on January 1st of `Year`**
* **`Female2`: Female population count for the `Age` group on December 31st of `Year`**
* **`Male2`: Male population count for the `Age` group on December 31st of `Year`**
* **`Total2`: Total population count for the `Age` group on December 31st of `Year`**
* **`OpenInterval`: Boolean indicator for whether or not the age interval is open to the right**

### `us_ppr`

**Parity progression ratios (PPRs) for baseline parity 0 through 3 for the U.S. 1918, 1943, and 1968 birth cohorts, respectively.**

* **`Cohort`: Calendar year of birth cohort**
* **`PPR[A]_[B]`: Parity progression ratio for moving from parity `A` to parity `B`**
* **`OpenInterval`: I'm guessing this is an indicator for whether or not the birth `Cohort` is still open; just assume that all of these cohorts have completed their reproductive lifespan**

### `us_fltper`, `us_mltper`, and `us_bltper`

**These are the annual female, male, and total population life table for five-year age groups for the U.S. from 1905 through 2016. They have all of the columns you expect from a life table, with `Age` being the starting age of each age group, and `Year` referring to your annual period from which the life table caluclated.**

# Annual age-specific person-year (ASPY) approximations {#aspy-approx}

## Computation

**Use the population counts provided to approximate for each sex (and for both sexes combined) the age-specific person-years (ASPY) lived during each individual calendar year period between 1 January 2005 and 1 January 2015, respectively, under the following assumptions:**

* **Linear (and constant) growth in person-years during the period**
* **Constant (but not necessarily linear) growth in person-years during the period**

```{r aspy_approx_compute}

```

## Report results

**Report total age-specific person-years lived in the 2015 calendar-year period below in a table with the following structure:**

* **`Age` as defined in the original dataset**
* **`py_linear`: The linear growth approximation of ASPY**
* **`py_constant`: The constant growth approximation of ASPY**
* **`py_hmd`: The ASPY approximation for the same population and period from the original HMD data source (i.e. `us_exposures`)**

```{r aspy_approx_report}

```


## Interpret results

**For the 2015 total population, compare these three age-specific ASPY approximations to one another. Then discuss the implications of these results for the linear and constant growth assumptions, including any variation by age group if applicable.**

**Answer below:**



# Annual crude growth rate (CGR) approximations by age and for the total population {#cgr_approx}

## Computation

**Using the person-year approximations you made in the [previous problem](#aspy-approx), estimate the annual CGR for each age and sex as well as for the total population for each calendar year period from 2005 through 2015 using all three person-year approximations, respectively:**

* **Linear (and constant) growth in person-years during the period**
* **Constant (but not necessarily linear) growth in person-years during the period**
* **Person-years as approximated by the HMD**

```{r cgr_approx_compute}

```

## Report results

### Age-specific CGR

**Report the age-specific CGR results for both sexes combined for 2015 in a table below with the following structure:**

* **`Age` as defined in the original dataset**
* **`cgr_linear`: age-specific CGR under the linear growth approximation of ASPY**
* **`cgr_constant`: age-specific CGR under the constant growth approximation of ASPY**
* **`cgr_hmd`: age-specific CGR using ASPY from the original HMD data source**

```{r cgr_approx_report_as}

```

### Population CGR

**Report the CGR results for the total population (both sexes combined) in a table below with the following structure:**

* **`Year`: The calendar-year period to which the CGR estimates apply**
* **`cgr_linear`: CGR under the linear growth approximation of ASPY**
* **`cgr_constant`: CGR under the constant growth approximation of ASPY**
* **`cgr_hmd`: CGR using ASPY from the original HMD data source**

```{r cgr_approx_report_total}

```

## Interpret results

**Describe the temporal pattern in CGR (as calculated from the HMD person-years approximations) from 2005 through 2015. Describe at least two alternative hypotheses for the pattern you see.**

**Answer below:**



# Principal period rates in 2015

## Computation

**For each sex and for both sexes combined, compute the principal period rates that contributed to the CGR (not age-specific) for each calendar-year period from 2005 through 2015. Hint: For each year, you will have to infer the crude rate of net migration (CRNM) because you can't calculate it directly.**

```{r principal_rates_compute}

```

## Report results

**Report your results for both sexes combined in a table below with the following structure:**

* **`Year`: The year to which the principal period rates apply**
* **`cbr/cdr/cnmr`: Separate columns for each of the principal period rates of interest**

## Interpret results

**Expand on your interpretation of the CGR results from the [previous problem](#cgr_approx) in light of the patterns you see in the principal period rates.**

**Answer below:**



# Age-standardized crude death rate (CDR) before, during, and recovering from the Great Recession

## Computation

**Using 2005 as a baseline standard, compute the age-standardized CDRs by sex (including both combined) for each of the calendar years from 2005 through 2015.**

```{r ascdr_compute}

```

## Report results

**Report your results in three tables below (one for males, one for females, one for both sexes), each with the following structure:**

* **`Year`: The year of the CGR estimate**
* **`cgr`: The un-standardized CGR**
* **`ascgr`: The standardized CGR**

```{r ascdr_report}

```

## Interpret results

**Within the context of the Great Recession, construct at least one hypothesis to explain the comparison between the temporal patterns in standardized vs. un-standardized CDR from 2005 through 2015.**

**Answer below:**



# CDR difference decomposition, 2015 vs. 2005

## Computation

**Decompose the difference in the un-standardized CDR between 2015 and 2005 (i.e. CDR(2015) - CDR(2005)) into its age-specific mortality schedule and age structure effects for each sex, respectively, include both combined. Also compute the mortality schedule and age structure effects for all ages combined (respectively for each sex, including both sexes combined).**

```{r cdr_decomp_compute}

```

## Report results

### Age-specific effects

**Report the age-specific decomposition results in three tables below (one for males, one for females, one for both sexes combined), each with the following structure:**

* **`Age`: The start year of the age group**
* **`m_effect`: The age-specific mortality schedule effect**
* **`c_effect`: The age-specific age structure effect**
* **`total_effect`: The total age-specific effect**

```{r cdr_decomp_report_as}

```

### Total effects

**Report the decomposition results for all ages combined in a table below with the following structure:**

* **`sex`: One row for Male, Female, and Total, respectively**
* **`m_effect`: The age-specific mortality schedule effect**
* **`c_effect`: The age-specific age structure effect**
* **`total_effect`: The total age-specific effect**

```{r cdr_decomp_report_total}

```

## Interpret results

**Within the context of the intervening Great Recession and its aftermath, interpret the age-specific and total decompositions of the change in CDR from 2005 to 2015. Make explicit reference to the interpretation of the signs and magnitudes of the effects themselves. Highlight any differences in the effects by sex that stand out to you.**

**Answer below:**



# Age-specific probabilities of death from Lexis triangles

## Computation

**For each sex (including all sexes combined), use the period birth counts (`us_births`) and the 2015 age-specific Lexis triangle death counts (`us_deaths_lexis`) to directly calculate the age-specific probabilities of death (i.e. ${}_nq_x$) for all age groups.**

```{r as_probs_compute}

```

## Report results

**Report the results below in a table with the following structure:**

* **`Age`: The starting age of the age group**
* **`qx_[m/f/b]`: Separate columns for the ${}_nq_x$ values of males (`m`), females (`f`) and both sexes combined (`b`)**

```{r as_probs_report}

```

# Cohort total Fertility Rate (TFR) from parity progression ratios (PPR)

## Computation

**Assuming (contrary to fact) that the maximum parity is 4, compute the cohort TFR from the PPRs in the U.S. in 1918, 1943, and 1968, respectively.**

```{r tfr_c_compute}

```

## Report results

**Report the results in a table below with the following structure:**

* **`Cohort`: The birth cohort to which the cohort TFR estimate applies**
* **`TFRc`: The estimated cohort TFR**

```{r tfr_c_report}

```

## What extra data do we need to calculate the cohort TFR correctly from PPRs?

**Answer below:**



## What would have to be true about parity above 4 for the cohort TFRs you calculated to be correct?

**Answer below:**



# PPR trend

**Below is a plot of the PPRs for parity 0 through 3 comparing 1918, 1943, and 1968.**

```{r ppr_plot, echo=FALSE, warning=FALSE, message=FALSE}
final_data[["us_ppr"]] %>%
  dplyr::select(-OpenInterval) %>%
  tidyr::pivot_longer(
    cols = PPR0_1:PPR3_4,
    names_to = "parity",
    names_prefix = "PPR",
    values_to = "ppr"
  ) %>%
  dplyr::mutate(
    parity = as.integer(substr(parity, 1, 1)),
    Cohort = factor(Cohort)
  ) %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = parity, y = ppr, color = Cohort) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid = element_blank())
```

## PPRs for 1918 vs. both later birth cohorts

**With reference to definition of a PPR and in terms of reproductive behavior (e.g. family planning, contraception, etc.), provide at least one potential explanation for the differences between the PPR curve of 1918 vs. both of the later cohorts.**

**Answer below:**



## PPRs for 1943 vs. 1968

**With reference to definition of a PPR and in terms of reproductive behavior (e.g. family planning, contraception, etc.), provide at least one potential explanation for the differences between the PPR curves of 1943 and 1968.**

**Answer below:**



# TFR, gross reproduction rate (GRR), net reproduction rate (NRR), generality fertility rate (GFR), and crude birth rate (CBR)

## Computation

**Calculate the TFR, GRR, NRR, and GFR for each of the years from 2005 through 2015. Yes, you can estimate each of these using the data provided.**

```{r fertility_and_reproduction_compute}

```

## Report results

**Report the results below in a table with the following structure:**

* **`Year`: The period for which the rate applies**
* **`[tfr/grr/nrr/gfr/cbr]`: Five separate columns for each of the period rates requested**

```{r fertility_and_reproduction_report}

```

## How well does the rule of thumb GFR = CBR/0.25  predict actual GFR?

**Answer below (you can compare the two simply by summarizing the central tendency and dispersion of their differences, if you like, but feel free to get creative):**



# Multiple decrement table for U.S. 2015: Deaths by despair vs. all other causes

## Computation

**Using the appropriate sex-specific parent life tables provided, construct sex-specific (including both sexes combined) multiple decrement tables for deaths by despair vs. all other causes.**

**Hints:**

* **You will need to recover the despair mortality rate and the rate of death due to all other causes from the despair death counts, total death counts, and all-cause mortality rates. You should definitely save the single decrement table columns for later.**
* **The despair death count's open-ended age interval is younger than the open-ended interval for the all-cause mortality data. Just assume that the death count ratio for the open-ended interval in the life table is the same as it is for open-ended interval in the cause-specific death count data.**

```{r multiple_decrement_compute}

```

## Report results

**For each sex (including both sexes combined), report a separate cause-specific multiple decrement table with column `Age`, as well as columns `mx_[d/o]`, `dx_[d/o]`, and `lx_[d/o]`, where `d` is for "despair" and `o` is for "other" causes of death, respectively. These column naming conventions should be familiar to you by now.**

```{r multiple_decrement_report}

```

## Interpret results

**Describe the age pattern of age-specific mortality of probability of death due to despair compared to deaths due to all other causes. Highlight any differences between the sexes that stand out to you.**

**Answer below:**



# Cause-deleted life table for U.S. 2015 without deaths by despair

## Compute

**Employing the proportional hazards assumption and accompanying procedure that we learned in class, construct a cause-deleted life table for each sex (including both sexes combined) that excludes mortality due to despair.**

```{r cause_deleted_compute}

```

## Report results

**For each sex (including both sexes combined), report a cause-deleted life table below with the typical life table columns that we've reported so many times in previous assignments (only they should be calculated for the cause-deleted case, of course).**

```{r cause_deleted_report}

```

## Interpret results

**With reference to the appropriate sex-specific parent life tables, compare the mortality experience with vs. without deaths due to despair in terms of: (1) differences in life expectancy; (2) differences in the probability of surviving from birth to age 30; and (3) differences in the probability of surviving from age 40 to age 60. Highlight any differences by sex that stand out to you.**

**Answer below**



# Population projection from 2005 to 2015

## Computation

**Assuming a closed population, project the female population of the U.S. by age from 2005 to 2015 assuming 2005 age- and sex-specific vital rates.**

**Hints:**

* **You will need to project over two five-year intervals!**
* **Your open-ended age interval for the projection needs to be five years younger than the open-ended age interval for your life tables, which means you'll need to aggregate the starting population counts accordingly.**
* **Your age groups need to be uniformly five-years in length, so again you need to aggregate the starting population counts accordingly.**
* **I have provided you with a female life table for the U.S. in 2005, but you will need to aggregate the 0-1 and 1-4 year age group's person-years columns into uniform five-year age groups.**

```{r projection_compute}

```

## Report results

**Report the results below in two tables with the following structure.**

* **`Age`: Starting age of the age interval**
* **`N_2005`: Age-specific starting population count on 1 Jan 2005**
* **`N_2015_proj`: Projected age-specific population count on 1 Jan 2015**
* **`N_2015_actual`: Actual age-specific population count on 1 Jan 2015**

```{r projection_report}

```

## Interpret results

**Compare the projected to the actual population in 2015. What could our projection model be missing that could explain any differences you see?**

**Answer below:**



# Stable equivalent population predictions with and without a COVID-19 baby bust

****

# Population momentum in 2019